Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-15 submission/manuscript.Rmd
edit figure legends

## Real time quantitative PCR (RT-qPCR)
nESCs were differentiated into EpiLCs as described above. Cells were lysed with Tri-reagent BD (Sigma #T3809) at 0, 24, and 48 hours of differentiation. RNA was phase separated with 0.1 uL/uL 1-bromo-3-chloropropane (Sigma #B9673) and then precipitated with with isopropanol (Sigma #I9516). For each sample, 2 ug of RNA was reverse transcribed using the ProtoScript II Reverse transcriptase kit from New England Biolabs (NEB #M0368S) and primed with oligo dT. Expression of _Kdm5c_ was detected using the primers 5'-CCCATGGAGGCCAGAGAATAAG-3' 5'-CTCAGCGGATAAGAGAATTTGCTAC-3' and normalized to TBP using Power SYBR™ Green PCR Master Mix (ThermoFisher #4367659).

@@

## Whole genome bisulfite sequencing (WGBS)
Genomic DNA (gDNA) from na&#239;ve ESCs and extended EpiLCs was extracted using the Wizard Genomic DNA Purification Kit (Promega A1120), following the instructions for Tissue Culture Cells. gDNA from two wild-types and two _Kdm5c_-KOs of each cell type was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150). Reads were adapter and quality trimmed with Trim Galore (v0.6.10) and aligned to the mm10 genome using Bismark (v0.22.1). Analysis of differential methylation at germline gene promoters was perfomed using methylKit (v1.28.0) with a minimum coverage of 3 paired reads, a percentage cut-off of 25%, and q-value of 0.01. Average percentage methylation at germline gene promoters was determined via methylKit (v1.28.0). Methylation bedgraph tracks were generated via Bismark and visualized using the UCSC genome browser.

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/Brain_tissue_genes.pdf}
  \caption[Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B). Left - MA plot of mRNA-sequencing. Right - Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal). Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red. \textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}. Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red. \textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}. Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/What_are_germline_genes.pdf}
  \caption[Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis (Mueller et al 2013). Expression is in Fragments Per Kilobase of transcript per Million mapped reads (FPKM). \textbf{C.} Number of testis DEGs that were classified as cell-type specific markers in a single cell RNA-seq dataset of the testis (Green et al 2018). Germline cell types are highlighted in green, somatic cell types in black. \textbf{D.} Sankey diagram of mouse genes filtered for germline enrichment based on their expression in wild-type and germline-depleted mice and in adult mouse non-gonadal tissues (Li et al 2017).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells. Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs). \textbf{B.} Representative images of wild-type (WT) and \textit{Kdm5c}-KO cells during ESC to EpiLC differentiation. Brightfield images taken at 20X. \textbf{C.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs. Welch's t-test, expression in transcripts per million (TPM). \textbf{D.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs. * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test. \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets. \textbf{F.} Average bigwigs of an example germline gene, \textit{D1Pas1}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple), amygdala (middle, red), and hippocampus (bottom, blue). \textbf{G.} enrichPlot gene ontology analysis comparing enriched biological processes for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs. \textbf{H.} Top left - Example germline identity DEGs unique to EpiLCs. Top right - Example 2-cell genes that are not dysregulated in \textit{Kdm5c}-KO EpiLCs. p-values for Welch's t-test. Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs.}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A.} Total number of germline-enriched RNA-seq DEGs for male hemizygous \textit{Kdm5c} knockout EpiLCs (XY 5CKO, purple), female heterozygous \textit{Kdm5c} knockout (XX 5CHET, orange), female homozygous \textit{Kdm5c} knockout (XX 5CKO, light orange) EpiLCs. \textbf{B.} Left - Eulerr overlap of \textit{Kdm5c} mutant male and female EpiLC germline DEGs. Right - Example of germline DEGs unique to females or males, \textit{Tktl2} and \textit{Esx1}.\textbf{C.} enrichPlot gene ontology analysis comparing enriched biological processes for germline DEGs shared between \textit{Kdm5c} mutant males and females (Shared), or unique to one sex (XX only or XY only). \textbf{D.} Heatmap of germline DEGs shared between male and female mutants. Color is the average log 2 fold change from sex-matched wild-type. \textbf{E.} Number of genes within each cluster from D. Clusters with higher expression in females compared to males (XX-biased) highlighted in orange. \textbf{F.} Example average bigwigs of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants. \textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome. Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant. X chromosome highlighted in black. \textbf{H.} Sankey diagram classifying egg-biased (pink) and sperm-biased (blue) and unbiased (purple) mouse germline-enriched genes. \textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants. \textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) germline genes dysregulated in both male and female \textit{Kdm5c} mutants.}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis.]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} \textbf{A.}  ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom). \textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS $\pm$ 500) in EpiLCs (purple) and PNCs (blue). \textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs. Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared). \textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right). \textbf{E.} Eulerr of number of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple). \textbf{F.} Example KDM5C ChIP-seq bigwigs of KDM5C binding at the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs. {G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs. \textbf{H.} Number of all gene promoters bound or unbound by KDM5C with instances of the E2F or E-box consensus sequence. \textbf{I.} HOMER motif analysis of all KDM5C-unbound germline gene promoters, highlighting significant enrichment of RFX family transcription factor motifs. \textbf{J.} Expression of RNA-seq DEG \textit{Rfx2} in wild-type and \textit{Kdm5c}-KO EpiLCs. P-value of Welch's t-test, expression in transcripts per million (TPM).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq peaks at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala. Right - Average H3K4me3 at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala. \textbf{B.} Left - Bigwigs of representative histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq peaks at representative germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs. Right - Average H3K4me2 at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs. \textbf{C.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation protocol and collection time points for RNA and protein. \textbf{D.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}). * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{E.} KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - representative lanes of Western blot for KDM5C and DAXX. * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters, based on UCSC annotation. Comparing all germline-enriched genes, KDM5C-bound germline genes, or KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of CGI-promoter versus non-CGI promoter germline genes. \textbf{H.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{I.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation at germline gene promoters (TSS $\pm$ 500) in wild-type ESCs versus exEpiLCs. Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink \textbf{J.} Volcano plot of WGBS of wild-type versus \textit{Kdm5c}-KO exEpiLCs. Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink, hypomethylated genes lackig a promoter CGI in blue. \textbf{K.} Example UCSC browser shots of germline genes of CpG methylation (CpGme) in wild-type and \textit{Kdm5c}-KO ESCs and exEpiLCs. \textbf{L.} Histogram of average percent CpGme at the promoter for germline genes with or without promoter CGIs. Wild-type in navy and \textit{Kdm5c}-KO in light blue. Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test.}
  \label{figurelabel}
\end{figure}

>>>>

## Real time quantitative PCR (RT-qPCR)
nESCs were differentiated into EpiLCs as described above. Cells were lysed with Tri-reagent BD (Sigma #T3809) at 0, 24, and 48 hours of differentiation. RNA was phase separated with 0.1 uL/uL 1-bromo-3-chloropropane (Sigma #B9673) and then precipitated with with isopropanol (Sigma #I9516). For each sample, 2 ug of RNA was reverse transcribed using the ProtoScript II Reverse transcriptase kit from New England Biolabs (NEB #M0368S) and primed with oligo dT. Expression of _Kdm5c_ was detected using the primers 5'-CCCATGGAGGCCAGAGAATAAG-3' 5'-CTCAGCGGATAAGAGAATTTGCTAC-3' and normalized to TBP with the Power SYBR™ Green PCR Master Mix (ThermoFisher #4367659).

@@

## Whole genome bisulfite sequencing (WGBS)
Genomic DNA (gDNA) from na&#239;ve ESCs and extended EpiLCs was extracted using the Wizard Genomic DNA Purification Kit (Promega A1120), following the instructions for Tissue Culture Cells. gDNA from two wild-types and two _Kdm5c_-KOs of each cell type was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150).  All samples had greater than 99% bisulfite conversion rates. Reads were adapter and quality trimmed with Trim Galore (v0.6.10) and aligned to the mm10 genome using Bismark (v0.22.1). Analysis of differential methylation at germline gene promoters was perfomed using methylKit (v1.28.0) with a minimum coverage of 3 paired reads, a percentage cut-off of 25%, and q-value of 0.01. Average percentage methylation at germline gene promoters was determined via methylKit (v1.28.0). Methylation bedgraph tracks were generated via Bismark and visualized using the UCSC genome browser.

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/Brain_tissue_genes.pdf}
  \caption[Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B). Left - MA plot of mRNA-sequencing. Right - Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - UCSC browser view of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal) (Average, n = 4). Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red. \textbf{D.} Left - UCSC browser view of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}. Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red. \textbf{E.} Left - UCSC browser view of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}. Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/What_are_germline_genes.pdf}
  \caption[Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis (Mueller et al 2013). Expression is in Fragments Per Kilobase of transcript per Million mapped reads (FPKM). \textbf{C.} Number of testis DEGs that were classified as cell-type specific markers in a single cell RNA-seq dataset of the testis (Green et al 2018). Germline cell types are highlighted in green, somatic cell types in black. \textbf{D.} Sankey diagram of mouse genes filtered for germline enrichment based on their expression in wild-type and W/Wv mice and in adult mouse non-gonadal tissues (Li et al 2017).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells. Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs). \textbf{B.} Representative images of male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) cells during ESC to EpiLC differentiation. Brightfield images taken at 20X. \textbf{C.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs. Welch's t-test, expression in transcripts per million (TPM). \textbf{D.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs. * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test. \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets. \textbf{F.} UCSC browser view of an example germline gene, \textit{D1Pas1}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple. Average, n = 3), amygdala (middle, red. Average, n = 4), and hippocampus (bottom, blue. Average, n = 4). \textbf{G.} enrichPlot gene ontology analysis comparing enriched biological processes for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs. \textbf{H.} Top left - Example germline identity DEGs unique to EpiLCs. Top right - Example 2-cell genes that are not dysregulated in \textit{Kdm5c}-KO EpiLCs. p-values for Welch's t-test. Bottom - UCSC browser view of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs (Average, n = 3).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A.} Total number of germline-enriched RNA-seq DEGs for male hemizygous \textit{Kdm5c} knockout EpiLCs (XY 5CKO, purple), female heterozygous \textit{Kdm5c} knockout (XX 5CHET, orange), female homozygous \textit{Kdm5c} knockout (XX 5CKO, light orange) EpiLCs. \textbf{B.} Left - Eulerr overlap of \textit{Kdm5c} mutant male and female EpiLC germline DEGs. Right - Example of germline DEGs unique to females or males, \textit{Tktl2} and \textit{Esx1}.\textbf{C.} enrichPlot gene ontology analysis comparing enriched biological processes for germline DEGs shared between \textit{Kdm5c} mutant males and females (Shared), or unique to one sex (XX only or XY only). \textbf{D.} Heatmap of germline DEGs shared between male and female mutants. Color is the average log 2 fold change from sex-matched wild-type. \textbf{E.} Number of genes within each cluster from D. Clusters with higher expression in females compared to males (XX-biased) highlighted in orange. \textbf{F.}  UCSC browser view of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants (Average, n = 3). \textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome. Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant. X chromosome highlighted in black. \textbf{H.} Sankey diagram classifying egg-biased (pink) and sperm-biased (blue) and unbiased (purple) mouse germline-enriched genes. \textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants. \textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) germline genes dysregulated in both male and female \textit{Kdm5c} mutants.}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis.]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} \textbf{A.}  ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom). \textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS $\pm$ 500) in EpiLCs (purple) and PNCs (blue). \textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs. Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared). \textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right). \textbf{E.} Eulerr overlap of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple). \textbf{F.} Example KDM5C ChIP-seq signal around the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs. \textbf{G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs. \textbf{H.} Number of all gene promoters bound or unbound by KDM5C with instances of the E2F or E-box consensus sequence. \textbf{I.} HOMER motif analysis of all KDM5C-unbound germline gene promoters, highlighting significant enrichment of RFX family transcription factor motifs. \textbf{J.} Expression of RNA-seq DEG \textit{Rfx2} in wild-type and \textit{Kdm5c}-KO EpiLCs. P-value of Welch's t-test, expression in transcripts per million (TPM).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} \textbf{A.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation and collection time points. \textbf{B.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs at 0, 24, and 48 hours of differentiation into EpiLCs.Expression calculated in comparision to TBP mRNA expression (2\textsuperscript{-deltaCT}). * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{C.} KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - representative lanes of Western blot for KDM5C and DAXX. * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{D.} Top - Representative UCSC browser view of histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq signal at two germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs. Bottom - Average H3K4me2 signal at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs. \textbf{E.} Top - Representative UCSC browser view of histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq signal at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala. Bottom - Average H3K4me3 signal at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala. \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters (CGI promoter, red), based on UCSC annotation. Left - percentage of all germline-enriched genes, middle - KDM5C-bound germline genes, and right - KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of germline genes with (CGI-promoter) or without (CGI-free) CGIs in their promoter. \textbf{H.} Expression of germline genes with (CGI-promoter, red) or without (CGI-free, salmon) CGIs in their promoter across stages of spermatogenesis from Green et al 2018. SPG - spermatogonia, Pre-Lep - preleptotene spermatocytes, S-Cytes - meiotic spermatocytes, S-Tids - post-meiotic haploid round spermatids, E.Sp - elongating spermatids. Wilcoxon test, * p < 0.05, ** p < 0.01, *** p < 0.001, **** p < 0.0001. \textbf{I.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{J.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation (CpGme) at germline gene promoters (TSS $\pm$ 500) in wild-type (WT) ESCs versus exEpiLCs. Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25%) with CGIs in red, CGI-free promoters in light blue \textbf{K.} Volcano plot of WGBS of wild-type (WT) versus \textit{Kdm5c}-KO exEpiLCs for germline gene promoters. Promoters with CGIs in red, CGI-free promoters in light blue. \textbf{L.} Histogram of average percent CpGme at the promoter of germline genes with or without CGIs. Wild-type in navy and \textit{Kdm5c}-KO (KO) in light blue. Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test. \textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs}
  \label{figurelabel}
\end{figure}
